---
title: "Calculate and compare repeatability for Ref and ALt alleles of the model mutations"
author: "Konstantin Popadin"
date: "29/07/2025"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list=ls(all=TRUE))
library(knitr)
# library(tidyverse)
# library(kableExtra)
library(dplyr)
library(tidyr)
library(ggplot2)
library(Biostrings)
library(stringdist) # install.packages("stringdist")
library(IRanges)

```
## Background
<br>
<br><br>

## 1. extract archives

```{r}
# Extract all CellKinetica tar.gz archives into an empty directory.
# Original archives are not deleted unless delete_archives = TRUE.

extract_tar_archives <- function(archive_dir, output_dir,
                                 archive_prefix = "repeatability_data",
                                 delete_archives = FALSE) {
  # Find archives in the correct order.
  archives <- list.files(
    archive_dir,
    pattern = paste0("^", archive_prefix, "_[0-9]+\\.tar\\.gz$"),
    full.names = TRUE
  )

  archives <- sort(archives)

  if (length(archives) == 0) {
    stop("No matching .tar.gz archives found in archive_dir.")
  }

  # Create output directory and check that it is empty.
  dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

  existing_files <- list.files(
    output_dir,
    recursive = TRUE,
    all.files = TRUE,
    include.dirs = FALSE
  )

  if (length(existing_files) > 0) {
    stop("The output directory is not empty. Please use an empty folder.")
  }

  # Extract archives one by one.
  for (archive in archives) {
    message("Extracting: ", basename(archive))

  utils::untar(
  tarfile = archive,
  exdir = output_dir,
  tar = "internal"
   )
  }

  # Confirm that files were extracted.
  extracted_files <- list.files(
    output_dir,
    recursive = TRUE,
    all.files = TRUE,
    include.dirs = FALSE
  )

  if (length(extracted_files) == 0) {
    stop("Extraction finished, but no files were found.")
  }

  message(
    "Extraction complete: ",
    length(extracted_files),
    " files restored."
  )

  # Optional deletion of archives after successful extraction.
  if (delete_archives) {
    file.remove(archives)
    message("Original archives deleted.")
  } else {
    message("Original archives were retained.")
  }

  invisible(extracted_files)
}


extract_tar_archives(
  archive_dir = "/home/popa/Repeatability/data/2_derived/01KP.ARCHIVE",
  output_dir = "/home/popa/Repeatability/data/2_derived/01KP.RESTORED",
  archive_prefix = "repeatability_data",
  delete_archives = FALSE
)

```
## Background
<br>
<br><br>

## 2. Loop to get the longest direct perfect repeats in the whole mtDNA 

# WHy? ToDO - take into account the circle? (I don't allow pos be on a margin of a repeat)
#Processed 1/66276 files (0.0%): 01KP.1.A.txt (no matches)
#Processed 2/66276 files (0.0%): 01KP.1.C.txt (no matches)
#Processed 3/66276 files (0.0%): 01KP.1.G.txt (no matches)
#Processed 4/66276 files (0.0%): 01KP.1.T.txt (no matches)

***
```{r}
# Paths
input_dir  <- "/home/popa/Repeatability/data/2_derived/01KP.RESTORED"
output_path <- "/home/popa/Repeatability/data/2_derived/01C.TheLongestPerfecInWholeMtDna.txt"

#major_arc_start <- 5798
#major_arc_end <- 16568

# List all .txt files in the folder
files <- list.files(input_dir, pattern = "\\.txt$", full.names = TRUE)
n_files <- length(files)

# Container for results
all_results <- list()

# Simple counter
i <- 0

for (f in files) {
  i <- i + 1
  
  # Read current file
  rep <- read.table(
    f,
    header = TRUE,
    colClasses = c(
      pos = "integer",
      nuc = "character",
      motif.seq = "character",
      motif.length = "integer",
      motif.start = "integer",
      motif.end = "integer",
      repeat.seq = "character",
      repeat.start = "integer",
      repeat.end = "integer",
      repeat.hamming.distance = "integer",
      effective.length = "integer"
    )
  )
  
  # Build logical conditions step by step
  motif_not_repeat <- !(rep$motif.start == rep$repeat.start & rep$motif.end == rep$repeat.end)
  #motif_in_major_arc <- rep$motif.start >= major_arc_start & rep$motif.end   <= major_arc_end
  #repeat_in_major_arc <- rep$repeat.start >= major_arc_start & rep$repeat.end   <= major_arc_end
  perfect_similarity <- rep$repeat.hamming.distance == 0
  pos_strictly_inside_motif <- rep$motif.start < rep$pos & rep$pos < rep$motif.end
  
  # idx <- motif_not_repeat & motif_in_major_arc & repeat_in_major_arc & perfect_similarity & pos_strictly_inside_motif # Combined index
  idx <- motif_not_repeat & perfect_similarity & pos_strictly_inside_motif # Combined index
  
  filtered_rep <- rep[idx, ] # Filter rows
  
  # Skip files with no matching rows
  if (nrow(filtered_rep) == 0) {
    cat(sprintf("Processed %d/%d files (%.1f%%): %s (no matches)\n",
                i, n_files, 100 * i / n_files, basename(f)))
    next
  }
  
  # Find maximal effective.length and keep all rows with that length
  max_len <- max(filtered_rep$effective.length, na.rm = TRUE)
  rep_max <- filtered_rep[filtered_rep$effective.length == max_len, ]
  
  # Store in list (no extra columns like 'file')
  all_results[[length(all_results) + 1]] <- rep_max
  
  cat(sprintf("Processed %d/%d files (%.1f%%): %s\n",
              i, n_files, 100 * i / n_files, basename(f)))
}

# Concatenate all results into one data frame
if (length(all_results) > 0) {
  final_df <- do.call(rbind, all_results)
  
  # Remove duplicated rows (keep one copy of each unique row)
  final_df <- unique(final_df)
  
  # Sort by pos (ascending)
  final_df <- final_df[order(final_df$pos), ]
  
  # Write to output file
  write.table(
    final_df,
    file = output_path,
    quote = FALSE,
    sep = "\t",
    row.names = FALSE
  )
} else {
  message("No rows matched the filters in any file.")
}

```
## Background
<br>
<br><br>

## 3: Loop to get the longest direct perfect repeats within the Majoir Arc

***
```{r}
# Paths
input_dir  <- "/home/popa/Repeatability/data/2_derived/01KP.RESTORED"
output_path <- "/home/popa/Repeatability/data/2_derived/01C.TheLongestPerfecWithinMajorArc.txt"

major_arc_start <- 5798
major_arc_end <- 16568

# List all .txt files in the folder
files <- list.files(input_dir, pattern = "\\.txt$", full.names = TRUE)
n_files <- length(files)

# Container for results
all_results <- list()

# Simple counter
i <- 0

for (f in files) {
  i <- i + 1
  
  # Read current file
  rep <- read.table(
    f,
    header = TRUE,
    colClasses = c(
      pos = "integer",
      nuc = "character",
      motif.seq = "character",
      motif.length = "integer",
      motif.start = "integer",
      motif.end = "integer",
      repeat.seq = "character",
      repeat.start = "integer",
      repeat.end = "integer",
      repeat.hamming.distance = "integer",
      effective.length = "integer"
    )
  )
  
  # Build logical conditions step by step
  motif_not_repeat <- !(rep$motif.start == rep$repeat.start & rep$motif.end == rep$repeat.end)
  motif_in_major_arc <- rep$motif.start >= major_arc_start & rep$motif.end   <= major_arc_end
  repeat_in_major_arc <- rep$repeat.start >= major_arc_start & rep$repeat.end   <= major_arc_end
  perfect_similarity <- rep$repeat.hamming.distance == 0
  pos_strictly_inside_motif <- rep$motif.start < rep$pos & rep$pos < rep$motif.end
  
  idx <- motif_not_repeat & motif_in_major_arc & repeat_in_major_arc & perfect_similarity & pos_strictly_inside_motif # Combined index
  # idx <- motif_not_repeat & perfect_similarity & pos_strictly_inside_motif # Combined index
  
  filtered_rep <- rep[idx, ] # Filter rows
  
  # Skip files with no matching rows
  if (nrow(filtered_rep) == 0) {
    cat(sprintf("Processed %d/%d files (%.1f%%): %s (no matches)\n",
                i, n_files, 100 * i / n_files, basename(f)))
    next
  }
  
  # Find maximal effective.length and keep all rows with that length
  max_len <- max(filtered_rep$effective.length, na.rm = TRUE)
  rep_max <- filtered_rep[filtered_rep$effective.length == max_len, ]
  
  # Store in list (no extra columns like 'file')
  all_results[[length(all_results) + 1]] <- rep_max
  
  cat(sprintf("Processed %d/%d files (%.1f%%): %s\n",
              i, n_files, 100 * i / n_files, basename(f)))
}

# Concatenate all results into one data frame
if (length(all_results) > 0) {
  final_df <- do.call(rbind, all_results)
  
  # Remove duplicated rows (keep one copy of each unique row)
  final_df <- unique(final_df)
  
  # Sort by pos (ascending)
  final_df <- final_df[order(final_df$pos), ]
  
  # Write to output file
  write.table(
    final_df,
    file = output_path,
    quote = FALSE,
    sep = "\t",
    row.names = FALSE
  )
} else {
  message("No rows matched the filters in any file.")
}

```
## Background
<br>
<br><br>

## 4: delete all txt files within the /home/popa/Repeatability/data/2_derived/01KP.RESTORED in order to push all to GitHub 
# all data stays in /home/popa/Repeatability/data/2_derived/01KP.ARCHIVE

***
```{r}
# Paths

# Delete all .txt files in the specified directory
output_dir <- "/home/popa/Repeatability/data/2_derived/01KP.RESTORED"

# List all .txt files (non-recursive, adjust if you need subfolders)
txt_files <- list.files(output_dir, pattern = "\\.txt$", full.names = TRUE)

# Remove them
if (length(txt_files) > 0) {
  file.remove(txt_files)
  cat("Deleted", length(txt_files), "files.\n")
} else {
  cat("No .txt files found.\n")
}

```

